The binary liquid model for gravity-driven microchannel simulations

### 2011-01-04 Text/paper.tex
Prepared according to the Chemical Engineering Journal. Michal, please indicate your address and do the last glimpse.

\documentclass{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{bm}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{color}
\usepackage{tabularx}

\title{The binary liquid model for gravity-driven microchannel simulations}
\date{\today}
\newcommand{\Ca}{\mathrm{Ca}}
\newcommand{\todo}[1]{{\color{red}#1}}
\begin{document}

\maketitle

\begin{abstract}
This work shows that the free-energy binary liquid lattice Boltzmann method is able to
reproduce the gravity driven microchannel flow of a gas bubble through liquid. In the regime of low
capillary
numbers
$\Ca$ (defined in Eq. \ref{capillary:number:definition}), the flow is thoroughly described by
Bretherton \cite{bretherton},
with the deposition film thickness proportional to $\Ca^{2/3}$. However, there are deviations
from the Bretherton correlation for $Ca\geq 0.005$, studied experimentally and numerically. This
work aims to examine the lattice Boltzmann binary liquid model feasibility to obtain valid results
for a bubble flow in liquid for a range of $Ca$ number. The binary
liquid model is a continuous-interface multiphase model allowing to properly resolve the interface.
It was found that the film interface needs to be as $50$
percent in comparison with the film thickness for simulation to give convergent results. While
the convergency criteria is fulfilled the results are of good comparison with the
numerical simulations of \citet{giavedoni-numerical} in the range of $Ca$ from $0.01$ to $1$. 
\end{abstract}

\section{Introduction}
The Taylor/Bretherton \cite{bretherton} flow describes the flow of long bubbles in
microchannels. It was found that the front meniscus film thickness is proportional
to $\Ca^{2/3}$ for capillary numbers less than $0.003$, where the capillary number is defined as:
\begin{equation}
\label{capillary:number:definition}
Ca=\frac{\mu_{liq} U_{bubble}}{\gamma},
\end{equation}
where $\mu_{liq}$ is the liquid viscosity, $U_{bubble}$ is a propagation bubble velocity, and
$\gamma$ is the surface tension between gas and liquid.
While there are number
of works which simulate the Bretherton problem with the boundary value method
\cite{ingham-plates,heil-bretherton}, those methods are of limited applicability for problems
involving complex geometries, free interface motion or for problems involving coalescence and/or
droplet breakup. The continuous interface models are more flexible for those kind of simulations.
However, if the interface is spread over several grid nodes, the question of
proper film thickness resolution in comparison with the interface arises.  This work
concentrates on resolving the parameter range of the binary liquid lattice
Boltzmann method to validate it in the range of capillary numbers.

The lattice Boltzmann method has emerged as a successful method to simulate
a wide variety of phenomena including hydrodynamics \cite{yu}, thermal flows
\cite{karlin-minimalmodels}, microflows \cite{ansumali-small-knudsen},
ferrofluids \cite{kuzmin-aniso}, and multiphase flows
\cite{swift,Shan-chen:extended}. Thanks to its kinetic nature, LBM as a particle
method easily tackles complex geometries andallows for incorporation of
physical phenomena on the microscopic level, as in the case in multiphase models. Most
multiphase lattice Boltzmann models \cite{swift, Shan-chen:extended} resolve
the interface in a continuous way. Such a representation
brings issues with film thickness resolution versus interface
resolution. In other words, when it comes to the resolution of the film
thickness the diffusive interface should be resolved in such a way as to have
negligible effect on the physics of the film.

This work examines the binary liquid free-energy LB model \cite{swift}. The
model simulates two liquids with the assumption of uniform overall
density. While the classical Bretherton problem is stated for gas and liquid, which are of largely
different densities and viscosities, it was indicated \cite{bretherton} that inertia effects were
neglected. Moreover, the results of \citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effect on the film thickness for the range of Reynolds numbers.
Therefore, the governing factor for microchannel flows is not the densities ratio, but the
viscosities ratio. Moreover, the results based on the uniform density, as in the case of the LBM
binary liquid model, are of good comparison with other simulations
\cite{giavedoni-numerical,heil-bretherton} taking into the account inertia effects.

The goal of this paper is to give an answer as to how to choose the
binary liquid model parameters, the viscosities ratio and the grid
resolution, in order to obtain correct film thicknesses.

The paper is organized as follows.  First, we briefly
explain the binary liquid lattice Boltzmann model and summarize the literature
for the Bretherton problem. Then, the parameters involved in the
simulation are examined and presented in the results section. The paper is
concluded with a summary of the main findings.

\section{Bretherton bubble flow}
Bretherton \cite{bretherton} studied a long bubble
moving in a tube filled with liquid. It was found that the deposition film thickness
is proportional to $\Ca^{2/3}$ in the range of small capillary numbers. Later on
it was realized \cite{wong-films,wong-pressure} that the film deposition
is proportional to $\Ca^{2/3}$ only in the specific region behind the front meniscus and
that the film thickness
varies over the bubble length for bubbles of finite length. Numerical
simulations \cite{giavedoni-numerical} and experimental studies
\cite{kreutzer-pressure-drop} showed as well a Reynolds number dependence, and
a deviation from the $\Ca^{2/3}$ rule for capillary numbers larger than $0.003$.
To be able to predict consistently the flow pattern for capillaries in
different ranges of parameters, simulations have been developed and are typically validated with the
small capillary numbers Bretherton problem.

There are a number of numerical methods which
were used for the simulation of the Taylor/Bretherton flow.
\citet{vanbaten-circular} studied the mass transfer and film
thickness for rising bubbles in a circular capillary using the finite volume method.
\citet{kreutzer-pressure-drop} used the finite volume method as well to perform
simulations of a circular capillary for a number of different
Reynolds and capillary numbers. \citeauthor{wong-films} \cite{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and showed the derivations of different shapes in the
slug crossections and menisci appearance.
\citet{heil-bretherton,ingham-plates} studied gas finger propagation in
a two-dimensional channel for a range of Reynolds and capillary
numbers using finite element method. \citet{giavedoni-numerical} performed crossvalidation of the
finite element solution developed by them with previously published results.
The solution is available for circular and planar cases.

While the applicability of other methods is well established for the
simulation of the Bretherton/Taylor problem, this is not the case for
the lattice Boltzmann method. A thorough parameter study for this approach
has not yet been done to the best of the authors' knowledge.

One should acknowledge the works of \citet{pagonabarraga-fingers} on menisci
in thin films for fingering phenomena. \citet{sehgal-microchannel} performed lattice Boltzmann
simulations of two-dimensional channel flows for larger capillary numbers, and
the authors found discrepancies with the classical Bretherton theory, which
is limited to the low capillary number regime \cite{giavedoni-numerical}.

To resolve the interface on a fine level for capillary numbers less than
$0.003$ is computationally intensive even in the two-dimensional case
(the amount of memory necessary to perform the simulation is on the order
of tens or hundreds of gigabytes).  This work is therefore based on a
comparison with the already established results, which were themselves
validated for the Bretherton bubble flow.  In this paper, we present
recipes for the initialization of the simulations and discuss optimal
parameter ranges for the binary liquid LB model.

\section{Lattice Boltzmann binary liquid model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another in certain
directions specified by the discrete velocity set.
The LBE is typically implementated as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{ collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $F_i$ is the external force population. The binary fluid LB model is
based on a free-energy functional \cite{swift,landau}, and operates with two
sets of populations: one to track the pressure and velocity fields, and the
other to represent the phase field.

The model we use in this work is a two-dimensional nine-velocity (D2Q9) model,
with equilibrium populations \cite{pooley-contact}:
\begin{equation}
\begin{aligned}
&f_i^{eq}&&=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr)\\
&&&+k\bigl(w_i^{xx} (\partial_x \phi)^2+w_i^{yy} (\partial_y \phi)^2 +w_i^{xy} \partial_x
\phi \partial_y \phi \bigr), i=1\div8\\
&f_0^{eq}&&=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}&&=w_i\left(\Gamma \mu + \frac{\phi c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi^m
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}\right), i=1\div8 \\
&g_0^{eq}&&=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; the chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is related to the surface
tension; $A$ is a parameter of the free-energy model  The bulk pressure
is expressed as $p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$. 
Parameters speficic to the D2Q9 grid are the weights
$w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and the tensor
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$ with
the sound speed parameter $c_s^2=1/3$.  The weights related to the
inclusion of the surface tension coefficient into the equations are as follows:
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$. The system can be shown to restore the macroscopic
fluid equations as:
\begin{equation}
\begin{aligned}
&\partial_t \rho+ \partial_{\alpha} \rho u_{\alpha}=0\\
&\rho\left(\partial_t+u_{\beta}\partial_{\beta}\right) u_{\alpha}=
-\partial_{\alpha}P_{\alpha \beta} +
\nu\partial_{\beta}\left(\partial_{\alpha}u_{\beta}+\partial_{\beta} u_{\alpha}\right)\\
&\partial_t \phi + \partial_{\alpha} \phi u_{\alpha}=M \partial^2_{\beta\beta} \mu,
\end{aligned}
\label{binary:fluid:system}
\end{equation}
where $\nu=c_s^2 (\tau-1/2)$ is the viscosity and
$M=\Gamma(\tau_{\phi}-1/2)$ is the mobility parameter. The system allows the separation of the liquid
phase with $\phi=1$ and a so-called gas phase with $\phi=-1$. The
relaxation time is taken as linearly dependant on the relaxation
times $\tau_{\mathrm{gas}}$ and $\tau_{\mathrm{liq}}$:
$\tau=\tau_{\mathrm{gas}}+\frac{\phi+1}{2}\tau_{\mathrm{liq}}$.
The surface tension in the framework of the binary liquid model is $\sqrt{\frac{8 k A}{9}}$.

\section{Lattice Boltzmann benchmark}
To properly simulate and to compare simulation results with the data
published in the literature, one needs to design a numerical benchmark addressing
the following challenges for the lattice Boltzmann method:
\begin{itemize}
 \item The diffusive nature of the interface in the multiphase model should not
influence the film thickness results. This is obviously achieved only for certain
resolutions of the interface in comparison with the film thickness. The same applies to spurious
currents arising from surface tension discretization which
can influence mass transfer in the thin film, thereby compromising the
overall solution. Therefore the diffusive interface should be of nengligible influence on the film
thickness.
 \item The wettability coefficient which in principle can control the
dynamic angle and the menisci appearance \cite{pagonabarraga-finger} should not
influence the overall solution.
 \item The free-energy model examined in the work is a single density, two
viscosities model. The classical Bretherton problem is a free-interface problem between liquid and
gas
where the film thickness is established by itself. Thus, the viscosity and density ratios for the
classical
Bretherton problem should
be large. However, in terms of the lattice Boltzmann binary liquid framework only the viscosities
ratio can be addressed. The results show that the results are consistent with other works and
inertia effects can be neglected.
  \item The Bretherton problem addresses a bubble flow driven by a pressure
difference, not a body force. However, to the best authors knowledge the pressure boundaries
conditions are still not yet developed for the binary liquid model, because it involves the
treatment of associated gradients and laplacians on the boundary.
This kind of pressure boundary conditions for the binary liquid model is currently
under development, and preliminary results show that the differences between
pressure difference flows and flows driven by a body force are small as far as the film
deposition thickness is of interest.  In this work we limit ourselves to the study
of body force driven flows, because of their simplicity and better numerical stability. The case of
body force driven flows implies the periodic boundary conditions. As soon as the periodic boundary
conditions are applied, not one bubble but whole bubble train is simulated. In this case one needs
to insure that bubbles do not influence each other making large distance between them.
 \item Most simulations are done for a channel with circular crosssection,
which is quite difficult to address in terms of lattice Boltzmann free energy
framework where one needs to calculate the Laplacian and gradients near boundaries.
Although one can apply certain finite difference stencils
\cite{arnold-boundary,hunt-boundary} developed specifically for curved boundaries,
the analysis of error introduced by the boundary conditions becomes an issue. Though the circular
axis-symmetric case can be simulated on rectangular grid by introducing specific mass and force
terms in the continuity and Navier-Stokes equation \cite{halliday-circular}, the lattice Boltzmann
incorporation of force and mass terms in the process of development and validation. Therefore
the benchmark is chosen as a plane-symmetric two-dimensional case.

\end{itemize}
Given all the concerns and challenges, the suggested lattice Boltzmann framework
is a two-dimensional flow driven by a body force.  An analytical
solution to this problem is to the best of our knowledge not yet known for the range of capillary
numbers.
Though some authors \cite{sehgal-microchannel} use the gravity driven model to
validate the Bretherton film thickness, the latter should be used with caution
since the velocity of the bubble is different from the fluid velocity and the
bubble shape
is different in the front and rear menisci depending on Bond number $Bo$. \citet{wong-films}
analytically described the film thickness variation along a bubble varying from $\Ca^{1/2}$ to
$\Ca^{2/3}$. The bubble should be long enough to have the film thickness proportional to
$Ca^{2/3}$. Therefore, we chose the bubble length to be $5$ channel heights. This number is shown
to be sufficient to give consistent with theory results. The
channel length is taken  as $3$ bubble lengths (or $15$ channel heights)
to minimize the influence of one bubble on another,
because of the periodicity of boundary conditions.
Figure \ref{fig:benchmark:sketch} is a sketch of the geometry
used in the benchmark.
\begin{figure}
%\includegraphics[width=0.97\textwidth]{}
\caption{{\color{red} Ask Sasha to help here. The benchmark sketch.}
\label{fig:benchmark:sketch}}
\end{figure}

\section{Results}
\subsection{The nondimensionalization and initialization procedure}
We now introduce the procedure of nondimensionalization, as well as
the initialization techniques we used to be able to determine parameters necessary for
simulation. The capillary number governs the interface thickness and is defined as:
\begin{equation}
Ca=\frac{U_{\mathrm{bubble}} \mu_{\mathrm{liq}}}{\gamma},
\end{equation}
where $\mu_{\mathrm{liq}}$ is the liquid viscosity, and $\gamma$ is the surface tension. The
bubble velocity $U_{\mathrm{bubble}}$ is taken in the center of the bubble where the film
thickness is examined.
While parameters such as $\mu_{\mathrm{liq}}$, $\gamma$ and the capillary number $\Ca$ are given
or defined through the model parameters, the bubble velocity $U_{\mathrm{bubble}}$ is not given
explicitly.
In the simulation we approximately match and initialize the velocity with the
Poiseuille profile center velocity, which is given by the equation:
\begin{equation}
\begin{aligned}
U_{\mathrm{bubble}} \approx U_{\mathrm{liq}}=\frac{H_{\mathrm{eff}}^2}{8
\mu_{\mathrm{liq}}}\frac{\mathrm{d}P}{\mathrm{d}x}\\
U_{\mathrm{bubble}}\approx
\frac{{(N_y-2)}^2}{\frac{8}{6}(\tau_{\mathrm{liq}}-\frac{1}{2})}\frac{\mathrm{d}P}{\mathrm{d}
x } ,
\end{aligned}
\end{equation}
where $N_y$ is the grid number in the vertical direction. As it will be discussed further, the
effective channel width is defined as $H_{\mathrm{eff}}=N_y-2$ because of bounce-back nodes.
By taking the Poiseuille profile one can estimate the body force through the
capillary number and grid resolution:
\begin{equation}
\label{poiseuille:velocity:center}
\begin{aligned}
&\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{{(N_y-2)}^2}\gamma Ca.
\end{aligned}
\end{equation}
The body force (\ref{poiseuille:velocity:center}) applied to the problem will cause the bubble
velocity different from the implied by the capillary number because of the multiphase flow. That's
why to achieve the 
desired velocity and the
capillary number the iteration procedure to find the appropriate body force needs to be done.
However, the initialization body force given by (\ref{poiseuille:velocity:center}) is a good
starting approximation point.

>>>>

%% This is file `elsarticle-template-1-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1-num.tex 149 2009-10-08 05:01:15Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1-num.tex $
%%
\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amsmath,amssymb}
\usepackage{bm}
%\usepackage[numbers,sort&compress]{natbib}
\usepackage{color}
\usepackage{tabularx}

%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
\usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

\journal{Chemical Engineering Journal}

\newcommand{\Ca}{\mathrm{Ca}}
\newcommand{\todo}[1]{{\color{red}#1}}

\begin{document}

\begin{frontmatter}
%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{The binary liquid model for gravity-driven microchannel simulations}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{shurik.kuzmin@gmail.com}
\author[google]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[uofa]{J.~Derksen}
\ead{kuzmin@ualberta.ca}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\address[google]{Google, Zurich, Switzerland}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[schlum]{DBR Schlumberger Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}

\begin{abstract}
This work shows that the free-energy binary liquid lattice Boltzmann method is able to
reproduce the gravity driven microchannel flow of a gas bubble through liquid. In the regime of low
capillary
numbers
$\Ca$ (defined in Eq. \ref{capillary:number:definition}), the flow is thoroughly described by
Bretherton \cite{bretherton},
with the deposition film thickness proportional to $\Ca^{2/3}$. However, there are deviations
from the Bretherton correlation for $Ca\geq 0.005$, studied experimentally and numerically. This
work aims to examine the lattice Boltzmann binary liquid model feasibility to obtain valid results
for a bubble flow in liquid for a range of $Ca$ number. The binary
liquid model is a continuous-interface multiphase model allowing to properly resolve the interface.
It was found that the interface thickness needs to be $50$
percent of the film thickness for simulation to give convergent results. While
the convergence criteria is fulfilled the results are in a good agreement with the
numerical simulations of \citet{giavedoni-numerical} in the range of $Ca$ from $0.01$ to $1$. 
\end{abstract}

\begin{keyword}
Bretherton problem \sep Microchannel simulation \sep Multiphase flow \sep Lattice Boltzmann method
\sep Binary liquid model \sep Flow between plates \sep Gravity driven
%% keywords here, in the form: keyword \sep keyword

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
\end{keyword}

\end{frontmatter}

%%
%% Start line numbering here if you want
%%
\linenumbers

%% main text

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \cite{key}          ==>>  [#]
%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
%%   \citet{key}         ==>>  Author [#]

\section{Introduction}
The Taylor/Bretherton \cite{bretherton} flow describes the flow of long bubbles in
microchannels. It was found that the front meniscus film thickness is proportional
to $\Ca^{2/3}$ for capillary numbers smaller than $0.003$, where the capillary number is defined as:
\begin{equation}
\label{capillary:number:definition}
Ca=\frac{\mu_{liq} U_{bubble}}{\gamma},
\end{equation}
where $\mu_{liq}$ is the liquid viscosity, $U_{bubble}$ is the bubble velocity, and
$\gamma$ is the surface tension between gas and liquid.

While there are number
of works which simulate the Bretherton problem by the boundary value approach
\cite{ingham-plates,heil-bretherton}, those methods are of limited applicability for problems
involving complex geometries, free interface motion, or for problems involving coalescence and/or
droplet breakup. The continuous interface models are more flexible for those kind of simulations.
However, if the interface is spread over several grid nodes, the question of
proper film resolution in comparison with the interface resolution arises.  This work is focused on
resolving the parameter range of the binary liquid lattice
Boltzmann method to validate it in the range of capillary numbers.

The lattice Boltzmann method has emerged as a successful method to simulate
a wide variety of phenomena including hydrodynamics \cite{yu}, thermal flows
\cite{karlin-minimalmodels}, microflows \cite{ansumali-small-knudsen},
ferrofluids \cite{kuzmin-aniso}, and multiphase flows
\cite{swift,Shan-chen:extended}. Thanks to its kinetic nature, LBM as a particle
method easily tackles complex geometries and allows incorporation of
physical phenomena on the microscopic level, as in the case of multiphase models. Most
multiphase lattice Boltzmann models \cite{swift, Shan-chen:extended} resolve
the interface using continuous interface methods where the interfaces spans over several grid nodes.
Such a representation
brings issues of the film thickness resolution versus interface
resolution. Thus, the diffusive interface should be resolved in such a way as to have
negligible effect on the physics of the film.

This work examines the binary liquid free-energy LB model \cite{swift}. The
model simulates two liquids with the assumption of uniform overall
density. While the classical Bretherton problem is stated for gas and liquid, which are of
significantly
different densities and viscosities, it was indicated \cite{bretherton} that inertia effects were
neglected. Moreover, the results of \citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effect on the film thickness for the relatively wide range of Reynolds
numbers.
Therefore, the major governing factor for microchannel flows is not the density ratio, but the
viscosity ratio. Moreover, the results based on the uniform density, as in the case of the LBM
binary liquid model, are good compared with other simulations
\cite{giavedoni-numerical,heil-bretherton} taking into the account inertia effects.

The goal of this paper is to understand how to choose the
binary liquid model parameters, the viscosity ratio and the grid
resolution to obtain correct film thicknesses.

The paper is organized as follows.  First, we briefly
explain the binary liquid lattice Boltzmann model and review the literature
for the Bretherton problem. Then, the parameters involved in the
simulation are examined and presented in the results section. The paper is
concluded with a summary of the main findings.

\section{Bretherton bubble flow}
Bretherton \cite{bretherton} studied a long bubble
moving in a tube filled with liquid. It was found that the deposition film thickness
is proportional to $\Ca^{2/3}$ in the range of small capillary numbers. Later on
it was realized \cite{wong-films,wong-pressure} that the film thickness
is proportional to $\Ca^{2/3}$ only in the certain region behind the front meniscus and
that the film thickness
varies over the bubble length for bubbles of finite length. Numerical
simulations \cite{giavedoni-numerical} and experimental studies
\cite{kreutzer-pressure-drop} showed a deviation from the $\Ca^{2/3}$ rule and Reynolds number 
effect on the film thickness for capillary numbers larger than $0.003$.
To  predict consistently a flow pattern for capillaries in
different ranges of parameters, simulations have been conducted and are typically validated with the
small capillary numbers Bretherton problem.

There are a number of numerical methods which
were used for the simulation of the Taylor/Bretherton flow.
\citet{vanbaten-circular} studied the mass transfer and film
thickness for rising bubbles in a circular capillary using the finite volume method.
\citet{kreutzer-pressure-drop} also used the finite volume method to perform
simulations of a circular capillary for a number of different
Reynolds and capillary numbers. \citeauthor{wong-films} \cite{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and showed calculated different shapes in the
slug cross sections and menisci appearance.
\citet{heil-bretherton,ingham-plates} studied gas finger propagation in
a two-dimensional channel for a range of Reynolds and capillary
numbers using finite element method. \citet{giavedoni-numerical} performed cross validation of the
finite element solution with previously published results.
The solutions were obtained for circular and planar cases.

While the applicability of other methods has been demonstrated for the
simulation of the Bretherton/Taylor problem, this is not the case for
the lattice Boltzmann method. A thorough parametric study for this approach
has not been done to the best of the authors' knowledge.

One should acknowledge the works of \citet{pagonabarraga-fingers} on menisci
in thin films for fingering phenomena. \citet{sehgal-microchannel} performed lattice Boltzmann
simulations of two-dimensional channel flows for relatively large capillary numbers, and
the authors found discrepancies with the classical Bretherton theory, which
is limited to the low capillary number regime \cite{giavedoni-numerical}.

Resolving the interface on a fine level for capillary numbers smaller than
$0.003$ is computationally expensive even in the two-dimensional case
(the amount of memory necessary to perform the simulation is of the order
of tens or hundreds gigabytes).  This work is therefore based on a
comparison of the results obtained with the known data
validated for the Bretherton bubble flow.  In this paper, we present
techniques for initialization of the simulations and discuss optimal
parameter ranges for the binary liquid LB model.

\section{Lattice Boltzmann binary liquid model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another in certain
directions specified by the discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{ collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$, $\omega$ is the
relaxation parameter, and $F_i$ is the external force population.

The binary fluid LB model is
based on a free-energy functional \cite{swift,landau}, and operates with two
sets of populations: one to track the pressure and the velocity fields, and another to represent the
phase field $\phi$ indicating the gas or liquid.

The model we use is a two-dimensional nine-velocity (D2Q9) model,
with equilibrium populations \cite{pooley-contact}:
\begin{equation}
\label{set:equilibrium:binary}
\begin{aligned}
&f_i^{eq}&&=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr)\\
&&&+k\bigl(w_i^{xx} (\partial_x \phi)^2+w_i^{yy} (\partial_y \phi)^2 +w_i^{xy} \partial_x
\phi \partial_y \phi \bigr), i=1\div8\\
&f_0^{eq}&&=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}&&=w_i\left(\Gamma \mu + \frac{\phi c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi^m
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}\right), i=1\div8 \\
&g_0^{eq}&&=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; the chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is the parameter related to the surface
tension; $A$ is the parameter of the free-energy model  The bulk pressure
is expressed as $p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$. 
Parameters specific to the D2Q9 grid are the weights
$w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and the tensor
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$ with
the sound speed parameter $c_s^2=1/3$.  The weights related to the
inclusion of the surface tension coefficient into the equations are as follows:
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$. The set of equations (\ref{set:equilibrium:binary}) restores the
macroscopic
fluid equations as:
\begin{equation}
\begin{aligned}
&\partial_t \rho+ \partial_{\alpha} \rho u_{\alpha}=0\\
&\rho\left(\partial_t+u_{\beta}\partial_{\beta}\right) u_{\alpha}=
-\partial_{\alpha}P_{\alpha \beta} +
\nu\partial_{\beta}\left(\partial_{\alpha}u_{\beta}+\partial_{\beta} u_{\alpha}\right)\\
&\partial_t \phi + \partial_{\alpha} \phi u_{\alpha}=M \partial^2_{\beta\beta} \mu,
\end{aligned}
\label{binary:fluid:system}
\end{equation}
where $\nu=c_s^2 (\tau-1/2)$ is the viscosity and
$M=\Gamma(\tau_{\phi}-1/2)$ is the mobility parameter.

The system allows the separation of the liquid
phase with $\phi=1$ and a so-called gas phase with $\phi=-1$. The
relaxation time is taken as linearly dependent on the relaxation
times $\tau_{\mathrm{gas}}$ and $\tau_{\mathrm{liq}}$:
$\tau=\tau_{\mathrm{gas}}+\frac{\phi+1}{2}\tau_{\mathrm{liq}}$.
The surface tension in the framework of the binary liquid model is $\sqrt{\frac{8 k A}{9}}$.
